---
title: 'Temporal Evolution of SARS-CoV-2 Vaccine Effectiveness: A Systematic Review and Meta-analysis'
Author: "Paddy Ssentongo"
output:
  html_document:
    df_print: paged
  pdf_document: default
  html_notebook: default
---

```{r, message=FALSE, echo=FALSE,warning=FALSE}

library(ggrepel)
library(colorspace)
library(colorblindr)
library(meta)
library(metafor)
library(splines)
library(dplyr)
library(ggsci)
```


```{r setup, include=FALSE,warning=FALSE,echo=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = FALSE,warning =F)
```




```{r,echo=FALSE,warning=FALSE}
# read data -----------------------------------------
rm(list=ls())
dat=read.csv("data/Vaccines6.csv")
```



```{r,echo=FALSE,warning=FALSE,fig.height=8, fig.width=15}

knots = quantile(dat$Time_day, p = c(0.25, 0.5, 0.75), na.rm = T)
mod_ns = lm(Estimate_any_sars_cov_2~ ns(Time_day, knots = knots), data = dat)

f <- dat %>%
   #filter(Country != "Qatar") %>%
  arrange(desc(N_vaccinated)) %>%
  ggplot(aes(x = Time_day, y = Estimate_any_sars_cov_2)) +
  #ggplot(aes(x = Period, y = Rates, fill=Period))+  
  #geom_smooth(se = F, method = "loess", size = 1.5,lty=2, col="orange") +
  #geom_smooth(method="gam", se=F,formula = y ~s(x),col="green")+
  #geom_smooth(method="gam", se=F, formula = y ~s(x, bs="cs"), col="black")+
  
  geom_smooth(method = "lm", formula = y ~ ns(x,knots = knots),col="red", lty=2, se=F)+
  #geom_smooth(method = "lm", formula = y ~ bs(x,knots = knots),col="blue", se=F) +
  #geom_boxplot(show.legend = F)+
  geom_point(aes(fill = Vaccine, size=N_vaccinated), shape = 21, alpha = .7) +
  expand_limits(x=c(0,200), y=c(0, 100))+
  #colorblindr::scale_fill_OkabeIto(name = "Vaccine type") +
  scale_fill_nejm(name = "Vaccine type")+
  scale_size(name = "N_vaccinated",
             range = c(2, 10), breaks = c(10000000, 500000000)) +
 # range = c(2, 10), breaks = c(10000, 500000, 1000000, 1500000, 2000000, 2500000, 4000000)) +
  #annotate("text", label = "p==0.33~~R^{2}==0.00", x = 120, y = 95, parse = T) +
  geom_text_repel(aes(label = Country), size = 2.5) +
  facet_wrap(~ "Vaccine effectiveness against SARS-CoV-2 Infection") +
  labs(x = "Time since full vaccination (days)", y = "VE (%)")  +
  geom_hline(yintercept=50, linetype="dashed", color = "blue", size=1.5)+
  guides(fill = guide_legend(override.aes = list(size = 3.5))) +
  theme_bw(base_size = 14) +
  theme(strip.text = element_text(face = "bold", size = 13))



```




### Effectiveness against symptomatic

```{r,echo=FALSE,warning=FALSE,fig.height=8, fig.width=15}
knots = quantile(dat$Time_day, p = c(0.25, 0.5, 0.75), na.rm = T)
f1 <- dat %>%
   #filter(Country != "Qatar") %>%
  arrange(desc(N_vaccinated)) %>%
  ggplot(aes(x = Time_day, y = Estimate_symptomatic)) +
  #ggplot(aes(x = Period, y = Rates, fill=Period))+  
  #geom_smooth(se = T, method = "loess", size = 1.5) +
  geom_smooth(method = "lm", formula = y ~ ns(x,knots = knots),col="red", lty=2, se=F)+
  #geom_boxplot(show.legend = F)+
  geom_point(aes(fill = Vaccine, size=N_vaccinated), shape = 21, alpha = .7) +
  expand_limits(x=c(0,200), y=c(0, 100))+
  #colorblindr::scale_fill_OkabeIto(name = "Vaccine type") +
  scale_fill_jama(name = "Vaccine type")+
  scale_size(name = "N_vaccinated",
             range = c(2, 10), breaks = c(10000000, 500000000)) +
 # range = c(2, 10), breaks = c(10000, 500000, 1000000, 1500000, 2000000, 2500000, 4000000)) +
  #annotate("text", label = "p==0.33~~R^{2}==0.00", x = 120, y = 95, parse = T) +
  geom_text_repel(aes(label = Country), size = 2.5) +
  facet_wrap(~ "Vaccine effectiveness against symptomatic") +
  geom_hline(yintercept=50, linetype="dashed", color = "blue", size=1.5)+
   labs(x = "Time since full vaccination (days)", y = "VE (%)")  +
  guides(fill = guide_legend(override.aes = list(size = 3.5))) +
  theme_bw(base_size = 14) +
  theme(strip.text = element_text(face = "bold", size = 13))


  

```




### Effectiveness against severe disease

```{r,echo=FALSE,warning=FALSE,fig.height=8, fig.width=15}

knots = quantile(dat$Time_day, p = c(0.25, 0.5, 0.75), na.rm = T)

f2 <- dat %>%
   #filter(Country != "Qatar") %>%
  arrange(desc(N_vaccinated)) %>%
  ggplot(aes(x = Time_day, y = Est_severe)) +
  geom_smooth(method = "lm", formula = y ~ ns(x,knots = knots),col="red", lty=2, se=F)+
  #ggplot(aes(x = Period, y = Rates, fill=Period))+  
  #geom_smooth(se = F, method = "loess", size = 1.5) +
  #geom_boxplot(show.legend = F)+
  geom_point(aes(fill = Vaccine, size=N_vaccinated), shape = 21, alpha = .7) +
   expand_limits(x=c(0,200), y=c(0, 100))+
  #colorblindr::scale_fill_OkabeIto(name = "Vaccine type") +
  scale_fill_jama(name = "Vaccine type")+
  scale_size(name = "N_vaccinated",
             range = c(2, 10), breaks = c(10000000, 500000000)) +
 # range = c(2, 10), breaks = c(10000, 500000, 1000000, 1500000, 2000000, 2500000, 4000000)) +
  #annotate("text", label = "p==0.33~~R^{2}==0.00", x = 120, y = 95, parse = T) +
  geom_text_repel(aes(label = Country), size = 2.5) +
  facet_wrap(~ "Vaccine effectiveness against severe COVID-19") +
  geom_hline(yintercept=50, linetype="dashed", color = "blue", size=1.5)+
  labs(x = "Time since full vaccination (days)", y = "VE (%)")  +
  guides(fill = guide_legend(override.aes = list(size = 3.5))) +
  theme_bw(base_size = 14) +
  theme(strip.text = element_text(face = "bold", size = 13))

```
# Figure 2

```{r, eval=T,echo=FALSE,warning=FALSE,fig.height=10, fig.width=20}

ggpubr::ggarrange(f, f1, f2, ncol = 1, common.legend = T,labels=c("A", "B","C"), legend = "top")
ggsave("figs7/Fig2.pdf", width = 15, height = 12)

```

### Effectiveness against asymptomatic

```{r,echo=FALSE,warning=FALSE,fig.height=5, fig.width=8}

knots = quantile(dat$Time_day, p = c(0.25, 0.5, 0.75), na.rm = T)
s10 <- dat %>%

   #filter(Country != "Qatar") %>%
  arrange(desc(N_vaccinated)) %>%
  ggplot(aes(x = Time_day, y = Estimate_asymptomatic)) +
  geom_smooth(method = "lm", formula = y ~ ns(x,knots = knots),col="red", lty=2, se=F)+
  #ggplot(aes(x = Period, y = Rates, fill=Period))+  
  #geom_smooth(se = T, method = "loess", size = 1.5) +
  #geom_boxplot(show.legend = F)+
  geom_point(aes(fill = Vaccine, size=N_vaccinated), shape = 21, alpha = .7) +
  expand_limits(x=c(0,200), y=c(0, 100))+
  #colorblindr::scale_fill_OkabeIto(name = "Vaccine type") +
  scale_fill_jama(name = "Vaccine type")+
  scale_size(name = "N_vaccinated",
             range = c(2, 10), breaks = c(10000000, 500000000)) +
 # range = c(2, 10), breaks = c(10000, 500000, 1000000, 1500000, 2000000, 2500000, 4000000)) +
  #annotate("text", label = "p==0.33~~R^{2}==0.00", x = 120, y = 95, parse = T) +
  geom_text_repel(aes(label = Country), size = 2.5) +
  facet_wrap(~ "VE against asymptomatic SARS-CoV-2 infection") +
  geom_hline(yintercept=50, linetype="dashed", color = "blue", size=1.5)+
    labs(x = "Time since full vaccination (days)", y = "VE (%)")  +
  guides(fill = guide_legend(override.aes = list(size = 3.5))) +
  theme_bw(base_size = 14) +
  theme(strip.text = element_text(face = "bold", size = 13))

ggsave("figs7/S10.pdf", width = 10, height = 8)  

```


# Age
# 65 and older infect

```{r,echo=FALSE,warning=FALSE,fig.height=8, fig.width=15}

library(splines)
knots = quantile(dat$Time_day, p = c(0.25, 0.5, 0.75), na.rm = T)

mod_ns = lm(X65infect~ ns(Time_day, knots = knots), data = dat)

fs1 <- dat %>%
   #filter(Country != "Qatar") %>%
  arrange(desc(N_vaccinated)) %>%
  ggplot(aes(x = Time_day, y =X65infect)) +
  #ggplot(aes(x = Period, y = Rates, fill=Period))+  
  #geom_smooth(se = F, method = "loess", size = 1.5,lty=2, col="orange") +
  #geom_smooth(method="gam", se=F,formula = y ~s(x),col="green")+
  #geom_smooth(method="gam", se=F, formula = y ~s(x, bs="cs"), col="black")+
  
  geom_smooth(method = "lm", formula = y ~ ns(x,knots = knots),col="red", lty=2, se=F)+
  #geom_smooth(method = "lm", formula = y ~ bs(x,knots = knots),col="blue", se=F) +
  #geom_boxplot(show.legend = F)+
  geom_point(aes(fill = Vaccine, size=N_vaccinated), shape = 21, alpha = .7) +
  expand_limits(x=c(0,200), y=c(0, 100))+
  #colorblindr::scale_fill_OkabeIto(name = "Vaccine type") +
  scale_fill_nejm(name = "Vaccine type")+
  scale_size(name = "N_vaccinated",
             range = c(2, 10), breaks = c(10000000, 500000000)) +
 # range = c(2, 10), breaks = c(10000, 500000, 1000000, 1500000, 2000000, 2500000, 4000000)) +
  #annotate("text", label = "p==0.33~~R^{2}==0.00", x = 120, y = 95, parse = T) +
  geom_text_repel(aes(label = Country), size = 2.5) +
  facet_wrap(~ "VE against SARS-CoV-2 Infection in 65 years and older") +
  labs(x = "Time since full vaccination (days)", y = "Estimated vaccine effectiveness (%)")  +
  geom_hline(yintercept=50, linetype="dashed", color = "blue", size=1.5)+
  guides(fill = guide_legend(override.aes = list(size = 3.5))) +
  theme_bw(base_size = 14) +
  theme(strip.text = element_text(face = "bold", size = 13))



```



# Age
# 65 years and older severe

```{r,echo=FALSE,warning=FALSE,fig.height=8, fig.width=15}
knots = quantile(dat$Time_day, p = c(0.25, 0.5, 0.75), na.rm = T)

mod_ns = lm(dat$X65severecritormort~ ns(Time_day, knots = knots), data = dat)

fs2 <- dat %>%
   #filter(Country != "Qatar") %>%
  arrange(desc(N_vaccinated)) %>%
  ggplot(aes(x = Time_day, y =X65severecritormort)) +
  #ggplot(aes(x = Period, y = Rates, fill=Period))+  
  #geom_smooth(se = F, method = "loess", size = 1.5,lty=2, col="orange") +
  #geom_smooth(method="gam", se=F,formula = y ~s(x),col="green")+
  #geom_smooth(method="gam", se=F, formula = y ~s(x, bs="cs"), col="black")+
  
  geom_smooth(method = "lm", formula = y ~ ns(x,knots = knots),col="red", lty=2, se=F)+
  #geom_smooth(method = "lm", formula = y ~ bs(x,knots = knots),col="blue", se=F) +
  #geom_boxplot(show.legend = F)+
  geom_point(aes(fill = Vaccine, size=N_vaccinated), shape = 21, alpha = .7) +
  expand_limits(x=c(0,200), y=c(0, 100))+
  #colorblindr::scale_fill_OkabeIto(name = "Vaccine type") +
  scale_fill_nejm(name = "Vaccine type")+
  scale_size(name = "N_vaccinated",
             range = c(2, 10), breaks = c(10000000, 500000000)) +
 # range = c(2, 10), breaks = c(10000, 500000, 1000000, 1500000, 2000000, 2500000, 4000000)) +
  #annotate("text", label = "p==0.33~~R^{2}==0.00", x = 120, y = 95, parse = T) +
  geom_text_repel(aes(label = Country), size = 2.5) +
  facet_wrap(~ "VE against severe disease in 65 years and older") +
  labs(x = "Time since full vaccination (days)", y = "Estimated vaccine effectiveness (%)")  +
  geom_hline(yintercept=50, linetype="dashed", color = "blue", size=1.5)+
  guides(fill = guide_legend(override.aes = list(size = 3.5))) +
  theme_bw(base_size = 14) +
  theme(strip.text = element_text(face = "bold", size = 13))


```




# Age
# less than 65 and older infection

```{r,echo=FALSE,warning=FALSE,fig.height=8, fig.width=15}

knots = quantile(dat$Time_day, p = c(0.25, 0.5, 0.75), na.rm = T)

mod_ns = lm(dat$lessthan65infect~ ns(Time_day, knots = knots), data = dat)

fs3 <- dat %>%
   #filter(Country != "Qatar") %>%
  arrange(desc(N_vaccinated)) %>%
  ggplot(aes(x = Time_day, y =lessthan65infect)) +
  #ggplot(aes(x = Period, y = Rates, fill=Period))+  
  #geom_smooth(se = F, method = "loess", size = 1.5,lty=2, col="orange") +
  #geom_smooth(method="gam", se=F,formula = y ~s(x),col="green")+
  #geom_smooth(method="gam", se=F, formula = y ~s(x, bs="cs"), col="black")+
  
  geom_smooth(method = "lm", formula = y ~ ns(x,knots = knots),col="red", lty=2, se=F)+
  #geom_smooth(method = "lm", formula = y ~ bs(x,knots = knots),col="blue", se=F) +
  #geom_boxplot(show.legend = F)+
  geom_point(aes(fill = Vaccine, size=N_vaccinated), shape = 21, alpha = .7) +
  expand_limits(x=c(0,200), y=c(0, 100))+
  #colorblindr::scale_fill_OkabeIto(name = "Vaccine type") +
  scale_fill_nejm(name = "Vaccine type")+
  scale_size(name = "N_vaccinated",
             range = c(2, 10), breaks = c(10000000, 500000000)) +
 # range = c(2, 10), breaks = c(10000, 500000, 1000000, 1500000, 2000000, 2500000, 4000000)) +
  #annotate("text", label = "p==0.33~~R^{2}==0.00", x = 120, y = 95, parse = T) +
  geom_text_repel(aes(label = Country), size = 2.5) +
  facet_wrap(~ "VE against SARS-CoV-2 Infection in less 65 years") +
  labs(x = "Time since full vaccination (days)", y = "Estimated vaccine effectiveness (%)")  +
  geom_hline(yintercept=50, linetype="dashed", color = "blue", size=1.5)+
  guides(fill = guide_legend(override.aes = list(size = 3.5))) +
  theme_bw(base_size = 14) +
  theme(strip.text = element_text(face = "bold", size = 13))



```




# Age
# less than 65 and older infection

```{r,echo=FALSE,warning=FALSE,fig.height=8, fig.width=15}

knots = quantile(dat$Time_day, p = c(0.25, 0.5, 0.75), na.rm = T)

mod_ns = lm(dat$lessthan65severecritormort~ ns(Time_day, knots = knots), data = dat)

fs4 <- dat %>%
   #filter(Country != "Qatar") %>%
  arrange(desc(N_vaccinated)) %>%
  ggplot(aes(x = Time_day, y =lessthan65severecritormort)) +
  #ggplot(aes(x = Period, y = Rates, fill=Period))+  
  #geom_smooth(se = F, method = "loess", size = 1.5,lty=2, col="orange") +
  #geom_smooth(method="gam", se=F,formula = y ~s(x),col="green")+
  #geom_smooth(method="gam", se=F, formula = y ~s(x, bs="cs"), col="black")+
  
  geom_smooth(method = "lm", formula = y ~ ns(x,knots = knots),col="red", lty=2, se=F)+
  #geom_smooth(method = "lm", formula = y ~ bs(x,knots = knots),col="blue", se=F) +
  #geom_boxplot(show.legend = F)+
  geom_point(aes(fill = Vaccine, size=N_vaccinated), shape = 21, alpha = .7) +
  expand_limits(x=c(0,200), y=c(0, 100))+
  #colorblindr::scale_fill_OkabeIto(name = "Vaccine type") +
  scale_fill_nejm(name = "Vaccine type")+
  scale_size(name = "N_vaccinated",
             range = c(2, 10), breaks = c(10000000, 500000000)) +
 # range = c(2, 10), breaks = c(10000, 500000, 1000000, 1500000, 2000000, 2500000, 4000000)) +
  #annotate("text", label = "p==0.33~~R^{2}==0.00", x = 120, y = 95, parse = T) +
  geom_text_repel(aes(label = Country), size = 2.5) +
  facet_wrap(~ "VE against severe disease in less 65 years") +
  labs(x = "Time since full vaccination (days)", y = "Estimated vaccine effectiveness (%)")  +
  geom_hline(yintercept=50, linetype="dashed", color = "blue", size=1.5)+
  guides(fill = guide_legend(override.aes = list(size = 3.5))) +
  theme_bw(base_size = 14) +
  theme(strip.text = element_text(face = "bold", size = 13))

ggpubr::ggarrange(fs3, fs1, fs4,fs2, ncol = 2, nrow=2, labels=c("A", "B", "C", "D"), common.legend = T, legend = "top")
ggsave("figs7/2Fig4.pdf", width = 15, height = 10)

```


## Part II

### Meta-analysis with Maximum Likelihood Method: Effectiveness against SARS-CoV-2 infections

```{r, eval=T,,echo=FALSE,warning=FALSE}

#dat=read.csv("data/VE_Clean_Final_meta.csv")
dat=read.csv("data/Vaccines6.csv")
dat <- dat[ which(dat$Month <6),]

l1=is.na(dat$Estimate_any_sars_cov_2a)
dat=dat[!l1,]

dat$mean=dat$Estimate_any_sars_cov_2a
#dat$seTE=(log(dat$RR_U-log(dat$RR_L))/(3.92))

dat$lower <- dat$LCI_any_sars_cov_2a

dat$upper <- dat$UCI_any_sars_cov_2a

dat$sd <- (dat$upper - dat$lower)/3.92
```



```{r, echo=FALSE,warning=FALSE, eval=T}
m<- metamean(n=N_vaccinated,
                   mean = mean,
                   sd = sd,
                   studlab = Study,
                    comb.fixed = FALSE,
                   comb.random = TRUE,
                   data = dat,
                   sm = "MRAW",
                   method.tau = "DL",
                   hakn =F,
                   title = "VE infections Overall")


```

# by month

```{r,fig.height=15, fig.width=10, echo=FALSE,warning=FALSE, eval=T}



mw<- update(m, byvar=Month, print.byvar=T)

pdf("figs7/Fig3.pdf", width = 10, height = 15)
forest(mw,
       #leftcols = c("Study"),
       leftlabs = c("Study"),
       digits = 2,
        bysort=T,
       overall= T, overall.hetstat = F,
       ref=NA,
       col.by="black",
       #sortvar = Year_Publication,
       rightcols=c("effect", "ci"),
       sortvar = mean,
       rightlabs = c("Mean VE", "95% CI"),
       squaresize=0.5,
       col.square="blue",
       col.diamond="maroon",
       col.diamond.lines="maroon",
       #xlab="Vaccine Effectiveness against SARS-CoV-2 infections (%)",
       xlab="Vaccine protection against SARS-CoV-2 infections (%)",
       print.pval.Q = T,
       digits.pval.Q=4,
       print.tau2=F,
       xlim = c(-20,100), at = c(-20, 0,20,40,60,80,100),pscale = 100,
       backtransf = T)
dev.off()

```


```{r, eval=T,,echo=FALSE,warning=FALSE}

dat=read.csv("data/Vaccines6.csv")


l1=is.na(dat$Estimate_any_sars_cov_2a)
dat=dat[!l1,]

dat$mean=dat$Estimate_any_sars_cov_2a
#dat$seTE=(log(dat$RR_U-log(dat$RR_L))/(3.92))

dat$lower <- dat$LCI_any_sars_cov_2a

dat$upper <- dat$UCI_any_sars_cov_2a

dat$sd <- (dat$upper - dat$lower)/3.92
```



```{r, echo=FALSE,warning=FALSE, eval=T}
m<- metamean(n=N_vaccinated,
                   mean = mean,
                   sd = sd,
                   studlab = Study,
                   comb.fixed = FALSE,
                   comb.random = TRUE,
                   data = dat,
                   sm = "MRAW",
                   method.tau = "DL",
                   hakn =F,
                   title = "VE infections Overall")


```



# Funnel plots : Risk ratio -----------------------------------------
```{r}

pdf("figs7/S11A.pdf", width = 15, height = 8)
funnel(m,xlab = "VE against SARS-CoV-2 infection (%)",studlab = F)
dev.off()

metabias(m, method="linreg")# Egger's linear regression test

```


```{r}

metabias(m, method="rank")

```


# by vaccine

```{r,fig.height=15, fig.width=10, echo=FALSE,warning=FALSE, eval=T}
mv<- update(m, byvar=Vaccine, print.byvar=T)


pdf("figs7/S1.pdf", width = 10, height = 12)
forest(mv,
       #leftcols = c("Study"),
       leftlabs = c("Study"),
       digits = 2,
        bysort=T,
       overall= T, overall.hetstat = F,
       ref=NA,
       col.by="black",
       #sortvar = Year_Publication,
       rightcols=c("effect", "ci"),
       sortvar = mean,
       rightlabs = c("Mean VE", "95% CI"),
       squaresize=0.5,
       col.square="blue",
       col.diamond="maroon",
       col.diamond.lines="maroon",
       xlab="Vaccine protection against SARS-CoV-2 infections (%)",
       print.pval.Q = T,
       digits.pval.Q=4,
       print.tau2=F,
       xlim = c(-20,100), at = c(-20, 0,20,40,60,80,100),pscale = 100,
       backtransf = T)
dev.off()

```




# by study design

```{r,fig.height=15, fig.width=10, echo=FALSE,warning=FALSE, eval=T}
mv<- update(m, byvar=RCT_or_Non.RCT, print.byvar=F)


pdf("figs7/S4.pdf", width = 10, height = 12)
forest(mv,
       #leftcols = c("Study"),
       leftlabs = c("Study"),
       digits = 2,
        bysort=T,
       overall= T, overall.hetstat = F,
       ref=NA,
       col.by="black",
       #sortvar = Year_Publication,
       rightcols=c("effect", "ci"),
       sortvar = mean,
       rightlabs = c("Mean VE", "95% CI"),
       squaresize=0.5,
       col.square="blue",
       col.diamond="maroon",
       col.diamond.lines="maroon",
       xlab="Vaccine protection against SARS-CoV-2 infections (%)",
       print.pval.Q = T,
       digits.pval.Q=4,
       print.tau2=F,
       xlim = c(-60,100), at = c(-60, -40, -20, 0,20,40,60,80,100),pscale = 100,
       backtransf = T)
dev.off()

```

# by country



```{r, eval=T,,echo=FALSE,warning=FALSE}

dat=read.csv("data/Vaccines6.csv")


l1=is.na(dat$Estimate_any_sars_cov_2a)
dat=dat[!l1,]

dat$mean=dat$Estimate_any_sars_cov_2a
#dat$seTE=(log(dat$RR_U-log(dat$RR_L))/(3.92))

dat$lower <- dat$LCI_any_sars_cov_2a

dat$upper <- dat$UCI_any_sars_cov_2a

dat$sd <- (dat$upper - dat$lower)/3.92
```



```{r, echo=FALSE,warning=FALSE, eval=T}
m<- metamean(n=N_vaccinated,
                   mean = mean,
                   sd = sd,
                   studlab = paste(Study, Month, sep = ", "),
                   comb.fixed = FALSE,
                   comb.random = TRUE,
                   data = dat,
                   sm = "MRAW",
                   method.tau = "DL",
                   hakn =F,
                   title = "VE infections Overall")


```


```{r,fig.height=15, fig.width=10, echo=FALSE,warning=FALSE, eval=T}
mv<- update(m, byvar=WHO, print.byvar=F)


pdf("figs7/S7.pdf", width = 10, height = 15)
forest(mv,
       #leftcols = c("Study"),
       leftlabs = c("Study, Time-Point"),
       digits = 2,
        bysort=T,
       overall= T, overall.hetstat = F,
       ref=NA,
       col.by="black",
       #sortvar = Year_Publication,
       rightcols=c("effect", "ci"),
       sortvar = mean,
       rightlabs = c("Mean VE", "95% CI"),
       squaresize=0.5,
       col.square="blue",
       col.diamond="maroon",
       col.diamond.lines="maroon",
       xlab="Vaccine protection against SARS-CoV-2 infections (%)",
       print.pval.Q = T,
       digits.pval.Q=4,
       print.tau2=F,
       xlim = c(-20,100), at = c(-20, 0,20,40,60,80,100),pscale = 100,
       backtransf = T)
dev.off()

```


### Age: Meta-analysis with DL: Effectiveness against SARS-CoV-2 infections: >65yrs

```{r, eval=T,,echo=FALSE,warning=FALSE}
dat=read.csv("data/Vaccines6.csv")

l1=is.na(dat$X65infect)
dat=dat[!l1,]

dat$mean=dat$X65infect
#dat$seTE=(log(dat$RR_U-log(dat$RR_L))/(3.92))

dat$lower <- dat$LCL_65infect

dat$upper <- dat$uCL_65infect

dat$sd <- (dat$upper - dat$lower)/3.92
```



```{r, echo=FALSE,warning=FALSE, eval=T}
m<- metamean(n=N_vaccinated,
                   mean = mean,
                   sd = sd,
                   studlab = Study,
                    comb.fixed = FALSE,
                   comb.random = TRUE,
                   data = dat,
                   sm = "MRAW",
                   method.tau = "DL",
                   hakn = T,
                   title = "VE infection >65")


```



```{r,fig.height=15, fig.width=10, echo=FALSE,warning=FALSE, eval=T}
#mw<- update(m, byvar=Month, print.byvar=T)

pdf("figs7/Figure4A.pdf", width = 10, height = 8)
forest(m,
       #leftcols = c("Study"),
       leftlabs = c("Study"),
       digits = 2,
        bysort=T,
       overall= T, overall.hetstat = F,
       ref=NA,
       col.by="black",
       #sortvar = Year_Publication,
       rightcols=c("effect", "ci"),
       sortvar = mean,
       rightlabs = c("Mean VE", "95% CI"),
       squaresize=0.5,
       col.square="blue",
       col.diamond="maroon",
       col.diamond.lines="maroon",
       xlab="Vaccine protection against SARS-CoV-2 infection: > 65 years (%)",
       print.pval.Q = T,
       digits.pval.Q=4,
       print.tau2=F,
       xlim = c(-20,100), at = c(-20, 0,20,40,60,80,100),pscale = 100,
       backtransf = T)
dev.off()

```


### Figure 4 


### Age: Meta-analysis with DL: Effectiveness against SARS-CoV-2 infections: <65yrs

```{r, eval=T,,echo=FALSE,warning=FALSE}
dat=read.csv("data/Vaccines6.csv")

l1=is.na(dat$lessthan65infect)
dat=dat[!l1,]

dat$mean=dat$lessthan65infect
#dat$seTE=(log(dat$RR_U-log(dat$RR_L))/(3.92))

dat$lower <- dat$lcllessthan65infect

dat$upper <- dat$ucllessthan65infect

dat$sd <- (dat$upper - dat$lower)/3.92
```



```{r, echo=FALSE,warning=FALSE, eval=T}
m<- metamean(n=N_vaccinated,
                   mean = mean,
                   sd = sd,
                   studlab = Study,
                    comb.fixed = FALSE,
                   comb.random = TRUE,
                   data = dat,
                   sm = "MRAW",
                   method.tau = "DL",
                   hakn = T,
                   title = "VE infect less than 65")


```



```{r,fig.height=15, fig.width=10, echo=FALSE,warning=FALSE, eval=T}
#mw<- update(m, byvar=Month, print.byvar=T)

pdf("figs7/Figure4B.pdf", width = 10, height = 8)
forest(m,
       #leftcols = c("Study"),
       leftlabs = c("Study"),
       digits = 2,
        bysort=T,
       overall= T, overall.hetstat = F,
       ref=NA,
       col.by="black",
       #sortvar = Year_Publication,
       rightcols=c("effect", "ci"),
       sortvar = mean,
       rightlabs = c("Mean VE", "95% CI"),
       squaresize=0.5,
       col.square="blue",
       col.diamond="maroon",
       col.diamond.lines="maroon",
       xlab="Vaccine protection against SARS-CoV-2 infections: < 65 years (%)",
       print.pval.Q = T,
       digits.pval.Q=4,
       print.tau2=F,
       xlim = c(-20,100), at = c(-20, 0,20,40,60,80,100),pscale = 100,
       backtransf = T)
dev.off()

```




### Age: Meta-analysis using DL method: Effectiveness against Severe COVID-19 Disease: >65

```{r,fig.height=6, fig.width=4, echo=FALSE,warning=FALSE, eval=T}

dat=read.csv("data/Vaccines6.csv")

l1=is.na(dat$X65severecritormort)
dat=dat[!l1,]

dat$mean=dat$X65severecritormort
#dat$seTE=(log(dat$RR_U-log(dat$RR_L))/(3.92))

dat$lower <- dat$LCL_65severecritormort

dat$upper <- dat$uCL_65severecritormort

dat$sd <- (dat$upper - dat$lower)/3.92
```



```{r, echo=FALSE,warning=FALSE, eval=T}
m1<- metamean(n=N_vaccinated,
                   mean = mean,
                   sd = sd,
                   studlab = Study,
                    comb.fixed = FALSE,
                   comb.random = TRUE,
                   data = dat,
                   sm = "MRAW",
                   method.tau = "DL",
                   hakn = T)


```



```{r,fig.height=15, fig.width=10, echo=FALSE,warning=FALSE, eval=T}
#mw1<- update(m1, byvar=Month, print.byvar=T)
pdf("figs7/Figure4C.pdf", width = 10, height = 8)
forest(m1,
       leftlabs = c("Study"),
       digits = 2,
        bysort=T,
       overall= T, overall.hetstat = F,
       ref=NA,
       col.by="black",
       #sortvar = Year_Publication,
       rightcols=c("effect", "ci"),
       sortvar = mean,
       rightlabs = c("Mean VE", "95% CI"),
       squaresize=0.5,
       col.square="blue",
       col.diamond="maroon",
       col.diamond.lines="maroon",
       xlab="Vaccine protection against severe COVID-19 disease: >65 years (%)",
       print.pval.Q = T,
       digits.pval.Q=4,
       print.tau2=F,
       xlim = c(-20,100), at = c(-20, 0,20,40,60,80,100),pscale = 100,
       backtransf = T)
dev.off()

```


### Age: Meta-analysis using DL method: Effectiveness against Severe COVID-19 Disease: < 65 years

```{r,fig.height=6, fig.width=4, echo=FALSE,warning=FALSE, eval=T}

dat=read.csv("data/Vaccines6.csv")

l1=is.na(dat$lessthan65severecritormort)
dat=dat[!l1,]

dat$mean=dat$lessthan65severecritormort
#dat$seTE=(log(dat$RR_U-log(dat$RR_L))/(3.92))

dat$lower <- dat$LCL_lessthan65severecritormort

dat$upper <- dat$uCL_lessthan65severecritormort

dat$sd <- (dat$upper - dat$lower)/3.92
```



```{r, echo=FALSE,warning=FALSE, eval=T}
m1<- metamean(n=N_vaccinated,
                   mean = mean,
                   sd = sd,
                   studlab = Study,
                    comb.fixed = FALSE,
                   comb.random = TRUE,
                   data = dat,
                   sm = "MRAW",
                   method.tau = "DL",
                   hakn = T)


```



```{r,fig.height=15, fig.width=10, echo=FALSE,warning=FALSE, eval=T}
#mw1<- update(m1, byvar=Month, print.byvar=T)
pdf("figs7/Figure4D.pdf", width = 10, height = 8)
forest(m1,
       leftlabs = c("Study"),
       digits = 2,
        bysort=T,
       overall= T, overall.hetstat = F,
       ref=NA,
       col.by="black",
       #sortvar = Year_Publication,
       rightcols=c("effect", "ci"),
       sortvar = mean,
       rightlabs = c("Mean VE", "95% CI"),
       squaresize=0.5,
       col.square="blue",
       col.diamond="maroon",
       col.diamond.lines="maroon",
       xlab="Vaccine protection against severe COVID-19 disease: < 65 years (%)",
       print.pval.Q = T,
       digits.pval.Q=4,
       print.tau2=F,
       xlim = c(-20,100), at = c(-20, 0,20,40,60,80,100),pscale = 100,
       backtransf = T)
dev.off()

```


### Figure 3............................................

### Meta-analysis using DL method: Effectiveness against Severe COVID-19 Disease

```{r,fig.height=6, fig.width=4, echo=FALSE,warning=FALSE, eval=T}

dat=read.csv("data/Vaccines6.csv")
dat <- dat[ which(dat$Month <6),]
#dat=read.csv("data/Vaccines5.csv")

l1=is.na(dat$Est_severea)
dat=dat[!l1,]


l1=is.na(dat$Hospitalization)
dat=dat[!l1,]

dat$mean=dat$Est_severea
#dat$seTE=(log(dat$RR_U-log(dat$RR_L))/(3.92))

dat$lower <- dat$lcl_severea

dat$upper <- dat$UCL_severea

dat$sd <- (dat$upper - dat$lower)/3.92
```



```{r, echo=FALSE,warning=FALSE, eval=T}
m1<- metamean(n=N_vaccinated,
                   mean = mean,
                   sd = sd,
                   studlab = Study,
                    comb.fixed = F,
                   comb.random = TRUE,
                   data = dat,
                   sm = "MRAW",
                   method.tau = "DL",
                   hakn = T)


```



# Funnel plots : Risk ratio -----------------------------------------
```{r}

pdf("figs7/S11C.pdf", width = 15, height = 8)
funnel(m1,xlab = "VE against Severe COVID-19  (%)",studlab = F)
dev.off()

metabias(m1, method="linreg")# Egger's linear regression test

```


$ month

```{r,fig.height=15, fig.width=10, echo=FALSE,warning=FALSE, eval=T}
mw1<- update(m1, byvar=Month, print.byvar=T)
#pdf("figs2/VE_Severe_ALL.pdf", width = 10, height = 15)
pdf("figs7/Fig6.pdf", width = 10, height = 12)
forest(mw1,
       leftlabs = c("Study"),
       digits = 2,
        bysort=T,
       overall= T, overall.hetstat = F,
       ref=NA,
       col.by="black",
       #sortvar = Year_Publication,
       rightcols=c("effect", "ci"),
       sortvar = mean,
       rightlabs = c("Mean VE", "95% CI"),
       squaresize=0.5,
       col.square="blue",
       col.diamond="maroon",
       col.diamond.lines="maroon",
      xlab="Vaccine protection against severe disease (%)",
       print.pval.Q = T,
       digits.pval.Q=4,
       print.tau2=F,
       xlim = c(-20,100), at = c(-20, 0,20,40,60,80,100),pscale = 100,
       backtransf = T)
dev.off()

```

# vaccine


```{r,fig.height=6, fig.width=4, echo=FALSE,warning=FALSE, eval=T}

dat=read.csv("data/Vaccines6.csv")

#dat=read.csv("data/Vaccines5.csv")

l1=is.na(dat$Est_severea)
dat=dat[!l1,]


l1=is.na(dat$Hospitalization)
dat=dat[!l1,]

dat$mean=dat$Est_severea
#dat$seTE=(log(dat$RR_U-log(dat$RR_L))/(3.92))

dat$lower <- dat$lcl_severea

dat$upper <- dat$UCL_severea

dat$sd <- (dat$upper - dat$lower)/3.92
```



```{r, echo=FALSE,warning=FALSE, eval=T}
m1<- metamean(n=N_vaccinated,
                   mean = mean,
                   sd = sd,
                   studlab = paste(Study, Month, sep = ", "),
                    comb.fixed = F,
                   comb.random = TRUE,
                   data = dat,
                   sm = "MRAW",
                   method.tau = "DL",
                   hakn = T)


```

```{r,fig.height=15, fig.width=10, echo=FALSE,warning=FALSE, eval=T}
mw1v<- update(m1, byvar=Vaccine, print.byvar=T)

pdf("figs7/S3.pdf", width = 10, height = 12)
forest(mw1v,
       leftlabs = c("Study, Time-point"),
       digits = 2,
        bysort=T,
       overall= T, overall.hetstat = F,
       ref=NA,
       col.by="black",
       #sortvar = Year_Publication,
       rightcols=c("effect", "ci"),
       sortvar = mean,
       rightlabs = c("Mean VE", "95% CI"),
       squaresize=0.5,
       col.square="blue",
       col.diamond="maroon",
       col.diamond.lines="maroon",
        xlab="Vaccine protection against severe disease (%)",
       print.pval.Q = T,
       digits.pval.Q=4,
       print.tau2=F,
       xlim = c(-20,100), at = c(-20, 0,20,40,60,80,100),pscale = 100,
       backtransf = T)
dev.off()

```

# by study design

```{r,fig.height=15, fig.width=10, echo=FALSE,warning=FALSE, eval=T}
mv1<- update(m1, byvar=RCT_or_Non.RCT, print.byvar=F)


pdf("figs7/S6.pdf", width = 10, height = 12)
forest(mv1,
       #leftcols = c("Study"),
       leftlabs = c("Study, Time-point"),
       digits = 2,
        bysort=T,
       overall= T, overall.hetstat = F,
       ref=NA,
       col.by="black",
       #sortvar = Year_Publication,
       rightcols=c("effect", "ci"),
       sortvar = mean,
       rightlabs = c("Mean VE", "95% CI"),
       squaresize=0.5,
       col.square="blue",
       col.diamond="maroon",
       col.diamond.lines="maroon",
       xlab="Vaccine protection against Severe Disease (%)",
       print.pval.Q = T,
       digits.pval.Q=4,
       print.tau2=F,
       xlim = c(-20,100), at = c(-20, 0,20,40,60,80,100),pscale = 100,
       backtransf = T)
dev.off()

```



# by country

```{r,fig.height=15, fig.width=10, echo=FALSE,warning=FALSE, eval=T}
mv1<- update(m1, byvar=WHO, print.byvar=F)


pdf("figs7/S9.pdf", width = 8, height = 12)
forest(mv1,
       #leftcols = c("Study"),
       leftlabs = c("Study, Time-point"),
       digits = 2,
        bysort=T,
       overall= T, overall.hetstat = F,
       ref=NA,
       col.by="black",
       #sortvar = Year_Publication,
       rightcols=c("effect", "ci"),
       sortvar = mean,
       rightlabs = c("Mean VE", "95% CI"),
       squaresize=0.5,
       col.square="blue",
       col.diamond="maroon",
       col.diamond.lines="maroon",
       xlab="Vaccine protection against severe disease (%)",
       print.pval.Q = T,
       digits.pval.Q=4,
       print.tau2=F,
       xlim = c(-20,100), at = c(-20, 0,20,40,60,80,100),pscale = 100,
       backtransf = T)
dev.off()

```


### Meta-analysis with Maximum Likelihood Method: Effectiveness against symptomatic infection

```{r, eval=T,,echo=FALSE,warning=FALSE}

dat=read.csv("data/Vaccines6.csv")
dat <- dat[ which(dat$Month <5),]

l1=is.na(dat$Estimate_symptomatic)
dat=dat[!l1,]

dat$mean=dat$Estimate_symptomatic
#dat$seTE=(log(dat$RR_U-log(dat$RR_L))/(3.92))

dat$lower <- dat$LCI_symptomatic

dat$upper <- dat$UCI_symptomatic

dat$sd <- (dat$upper - dat$lower)/3.92
```



```{r, echo=FALSE,warning=FALSE, eval=T}
m<- metamean(n=N_vaccinated,
                   mean = mean,
                   sd = sd,
                   studlab = Study,
                    comb.fixed = FALSE,
                   comb.random = TRUE,
                   data = dat,
                   sm = "MRAW",
                   method.tau = "DL",
                   hakn =F,
                   title = "VE infections Overall")


```


# Funnel plots  -----------------------------------------
```{r}

pdf("figs7/S11B.pdf", width = 15, height = 8)
funnel(m,xlab = "VE against symptomatic COVID-19  (%)",studlab = F)
dev.off()

metabias(m, method="linreg")# Egger's linear regression test

```


# by month

```{r,fig.height=15, fig.width=10, echo=FALSE,warning=FALSE, eval=T}

mw<- update(m, byvar=Month, print.byvar=T)

pdf("figs7/Fig5.pdf", width = 10, height = 10)
forest(mw,
       #leftcols = c("Study"),
       leftlabs = c("Study"),
       digits = 2,
        bysort=T,
       overall= T, overall.hetstat = F,
       ref=NA,
       col.by="black",
       #sortvar = Year_Publication,
       rightcols=c("effect", "ci"),
       sortvar = mean,
       rightlabs = c("Mean VE", "95% CI"),
       squaresize=0.5,
       col.square="blue",
       col.diamond="maroon",
       col.diamond.lines="maroon",
       #xlab="Vaccine Effectiveness against SARS-CoV-2 infections (%)",
       xlab="Vaccine protection against symptomatic infection (%)",
       print.pval.Q = T,
       digits.pval.Q=4,
       print.tau2=F,
       xlim = c(-20,100), at = c(-20, 0,20,40,60,80,100),pscale = 100,
       backtransf = T)
dev.off()

```


```{r, eval=T,,echo=FALSE,warning=FALSE}

dat=read.csv("data/Vaccines6.csv")
dat <- dat[ which(dat$Month <5),]

l1=is.na(dat$Estimate_symptomatic)
dat=dat[!l1,]

dat$mean=dat$Estimate_symptomatic
#dat$seTE=(log(dat$RR_U-log(dat$RR_L))/(3.92))

dat$lower <- dat$LCI_symptomatic

dat$upper <- dat$UCI_symptomatic

dat$sd <- (dat$upper - dat$lower)/3.92
```



```{r, echo=FALSE,warning=FALSE, eval=T}
m<- metamean(n=N_vaccinated,
                   mean = mean,
                   sd = sd,
                   studlab = paste(Study, Month, sep = ", "),
                    comb.fixed = FALSE,
                   comb.random = TRUE,
                   data = dat,
                   sm = "MRAW",
                   method.tau = "DL",
                   hakn =F,
                   title = "VE infections Overall")


```






# vaccine

```{r,fig.height=15, fig.width=10, echo=FALSE,warning=FALSE, eval=T}
mw1v<- update(m, byvar=Vaccine, print.byvar=T)

pdf("figs7/S2.pdf", width = 10, height = 12)
forest(mw1v,
       leftlabs = c("Study, Time-point"),
       digits = 2,
        bysort=T,
       overall= T, overall.hetstat = F,
       ref=NA,
       col.by="black",
       #sortvar = Year_Publication,
       rightcols=c("effect", "ci"),
       sortvar = mean,
       rightlabs = c("Mean VE", "95% CI"),
       squaresize=0.5,
       col.square="blue",
       col.diamond="maroon",
       col.diamond.lines="maroon",
        xlab="Vaccine protection against symptomatic infection (%)",
       print.pval.Q = T,
       digits.pval.Q=4,
       print.tau2=F,
       xlim = c(-20,100), at = c(-20, 0,20,40,60,80,100),pscale = 100,
       backtransf = T)
dev.off()

```




# by study design




```{r,fig.height=15, fig.width=10, echo=FALSE,warning=FALSE, eval=T}
mv1<- update(m, byvar=RCT_or_Non.RCT, print.byvar=F)


pdf("figs7/S5.pdf", width = 10, height = 12)
forest(mv1,
       #leftcols = c("Study"),
       leftlabs = c("Study, Time-point"),
       digits = 2,
        bysort=T,
       overall= T, overall.hetstat = F,
       ref=NA,
       col.by="black",
       #sortvar = Year_Publication,
       rightcols=c("effect", "ci"),
       sortvar = mean,
       rightlabs = c("Mean VE", "95% CI"),
       squaresize=0.5,
       col.square="blue",
       col.diamond="maroon",
       col.diamond.lines="maroon",
       xlab="Vaccine protection against symptomatic infection (%)",
       print.pval.Q = T,
       digits.pval.Q=4,
       print.tau2=F,
       xlim = c(-20,100), at = c(-20, 0,20,40,60,80,100),pscale = 100,
       backtransf = T)
dev.off()

```




# by country

```{r, eval=T,,echo=FALSE,warning=FALSE}

dat=read.csv("data/Vaccines6.csv")
#dat <- dat[ which(dat$Month <5),]

l1=is.na(dat$Estimate_symptomatic)
dat=dat[!l1,]

dat$mean=dat$Estimate_symptomatic
#dat$seTE=(log(dat$RR_U-log(dat$RR_L))/(3.92))

dat$lower <- dat$LCI_symptomatic

dat$upper <- dat$UCI_symptomatic

dat$sd <- (dat$upper - dat$lower)/3.92
```



```{r, echo=FALSE,warning=FALSE, eval=T}
m<- metamean(n=N_vaccinated,
                   mean = mean,
                   sd = sd,
                   studlab = paste(Study, Month, sep = ", "),
                    comb.fixed = FALSE,
                   comb.random = TRUE,
                   data = dat,
                   sm = "MRAW",
                   method.tau = "DL",
                   hakn =F,
                   title = "VE infections Overall")


```


```{r,fig.height=15, fig.width=10, echo=FALSE,warning=FALSE, eval=T}
mv1<- update(m, byvar=WHO, print.byvar=F)


pdf("figs7/S8.pdf", width = 8, height = 12)
forest(mv1,
       #leftcols = c("Study"),
       leftlabs = c("Study, Time-point"),
       digits = 2,
        bysort=T,
       overall= T, overall.hetstat = F,
       ref=NA,
       col.by="black",
       #sortvar = Year_Publication,
       rightcols=c("effect", "ci"),
       sortvar = mean,
       rightlabs = c("Mean VE", "95% CI"),
       squaresize=0.5,
       col.square="blue",
       col.diamond="maroon",
       col.diamond.lines="maroon",
       xlab="Vaccine protection against symptomatic infection (%)",
       print.pval.Q = T,
       digits.pval.Q=4,
       print.tau2=F,
       xlim = c(-20,100), at = c(-20, 0,20,40,60,80,100),pscale = 100,
       backtransf = T)
dev.off()

```




